##Introduction treebase package in R.

####Emily Woodruff  April 17, 2015

-----------------------------------------
#### What is treebase?

Treebase is an R package written to access trees from TreeBASE. It was written by Carl Bottiger and Duncan Templ Lang in 2012, see their [paper](http://onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2012.00247.x/full) published in _Methods in Ecology and Evolution_.

[TreeBASE](http://treebase.org/treebase-web/urlAPI.html) is an online repository of published phylogenetic trees and data.  

#### Why is it useful? 

- Acquire many trees at once from TreeBASE repository
- Sort through trees in TreeBASE by keyword, author, etc.
- Download metadata from TreeBASE or [Dryad](http://datadryad.org/), another free online data repository for published scientific and medical data, not just trees!  

###Getting started with treebase
1. `treebase` dependencies: `ape`

`treebase` is available [here](https://github.com/ropensci/treebase) on github or you can install directly from R:

```{r installing, echo=TRUE, eval=FALSE}
install.packages("treebase", repos="https://github.com/ropensci/treebase" )
```
2. Load 'treebase' library
```{r loading treebase, eval=TRUE}
library(treebase)
```

###Functions within treebase

###1. **search treebase**  

#####Use this to search for trees and data matrices on TreeBASE

#####Commands for _search_ _treebase_ function (default settings)
```{r search commands, echo=TRUE, eval=FALSE}
trees <- search_treebase("keyword", by="subject", returns = c("tree", "matrix"), exact_match = FALSE, max_trees = Inf, branch_lengths = FALSE, curl = getCurlHandle(), verbose = TRUE,  pause1 = 2, pause2 = 1, attempts = 3, only_metadata = FALSE) 
```

#####Key arguments for _search_ _treebase_:
**input**   search query  
**by**      search type:

- keywords
- author
- subject
- doi or ncbi accession
- kind.tree
- type.tree
- ntax
- quality
- study
- taxon

**returns**        do you want a tree? or a character matrix?
**exact_match**    forces exact matching, otherwise does partial match
**max_trees**      max trees to return
**branch_lengths** returns branch lengths too
**only_metadata**  returns only metadata, not phylogenies
**curl**           handles repeated calls to server


#####To search for trees by an author
```{r author search, echo=TRUE, eval=TRUE}
tree <- search_treebase("Ronquist", by=c("author"), returns= c("tree"), exact_match=TRUE, max_trees = 5)[[1]]
```

Without [[1]], `tree` will refer to ALL the trees you just pulled in, not just the first tree:
```{r author search2, echo=TRUE, eval=FALSE}
tree <- search_treebase("Ronquist", by=c("author"), returns= c("tree"), exact_match=TRUE, max_trees = 5)
```

Note: searches may take a while, depending on your query

#####Look at first 3 trees pulled in:
```{r look at trees, echo=TRUE, eval=FALSE}
tree[1:3]
#Let's check out the structure of the data we pulled down from TreeBASE
str(tree)
```

#####Let's look at the first tree now:
```{r plot first tree, echo=TRUE, eval=TRUE}
library(ape)
plot(tree)
# Making it a little easier to read...
plot(tree, type="phylogram", edge.width=1.5, cex=0.5)
```

#####To gather metadata only (not phylogenies) using _search_ _trees_ function:
```{r metadata only, echo=TRUE, eval=TRUE}
tree.data <- search_treebase("Ronquist", by=c("author"), returns= c("tree"), exact_match=TRUE, max_trees = 2, only_metadata=TRUE)
str(tree.data[1:2])
## It's a list! Making it a data frame...
tree.data2 <- do.call(rbind, lapply(tree.data, data.frame, stringsAsFactors=FALSE))
## Looking at our new data frame
head(tree.data2)
```
#####Now let's plot some of the metadata
```{r plot metadata, echo=TRUE, eval=TRUE}
plot(tree.data2$ntax, main="Number of taxa in each study", xlab="Study", ylab="N taxa", pch=16, col="blue")
```

This is a little bit boring.

#####Let's make a barplot in ggplot
```{r barplot, echo=TRUE, eval=TRUE}
library(ggplot2)
ggplot(tree.data2, aes(x=S.id, y=ntax, color=S.id)) + 
  geom_point(size=3.5, alpha = 1, position = position_jitter())
```

#####Now let's look at a barplot of the number of taxa in each study, colored by study ID.
#####Note there are multiple trees from the same study
```{r geom bar, echo=TRUE, eval=TRUE}
ggplot(tree.data2, aes(ntax, color=S.id)) + 
  geom_bar() + 
  geom_histogram(binwidth=0.5)
```

###2. **download metadata**

#####Use this to download metadata from TreeBASE

#####Commands for _download_ _metadata_ function
```{r metadata commands, echo=TRUE, eval=FALSE}
download_metadata(query = "", by = c("all", "until", "from"), curl = getCurlHandle())
```
#####Key arguments for _download_ _metadata_: 

**date** yyyy-mm-dd 
**by** return data from specified date "until", "from" or "all" 
**curl** 

#####Getting metadata for all submissions dated from 1980 until 2013:
```{r metadata until, echo=TRUE, eval=TRUE}
data <- download_metadata("2013-01-01", by="until")
pubdate <- sapply(data, function(x) as.numeric(x$date))
hist(pubdate, main="Publication Dates", xlab="Year", col="green")
```

#####Getting metadata for all submissions from 2013 to the present:
```{r metadata from, echo=TRUE, eval=TRUE}
data <- download_metadata("2013-01-01", by="from")
#Note: data pulled down is a list
pubdate <- sapply(data, function(x) as.numeric(x$date))
hist(pubdate, main="Publication Dates", xlab="Year", col="green")
```

###3. **dryad metadata**

#####Use this to download metadata from specific studies from Dryad

#####Commands for _dryad_ _metadata_ function
```{r dryad commands, echo=TRUE, eval=FALSE}
dryad_metadata(study.id, curl = getCurlHandle())
```

#####Key arguments for _dryada_ _metadata_:
**study.id**    Dryad study ID
```{r dryad example, echo=TRUE, eval=TRUE}
metadata <- dryad_metadata("10255/dryad.12")
```
Note: Dryad is phasing out the handles used to identify older submissions, now studies are identified exclusively using DOIs. I tried this function with a newer DOI but I couldn't get it to work...

###4. **cache treebase**

#####Use this to store phylogenies from TreeBASE locally

#####Commands for _cache_ _treebase_ function
```{r cache commands, echo=TRUE, eval=FALSE}
cache_treebase(file=paste("treebase-"), Sys,Date(), ".rda", sep=""),
pause1 = 3, pause2 = 3, attempts = 10, max_trees = Inf, only_metadata = FALSE, save = TRUE)
```
Note: this can take a long time! Recommended as an overnight run :)

